## Exploratory Data Analysis

\(\ \)

```{r}
n <- 864
df <- read.csv(file="pollutants.csv")
# the X variables is the same as the index so we can delete it
df = df[,-1]
# since the five kinds of cells other than white blood cells
# are all recorded in percentage, out of white blood cells,
# we first multiply all five by the count of white blood cells
# to get the count of each.
df[21] = df[20] * df[21] / 100
df[22] = df[20] * df[22] / 100
df[23] = df[20] * df[23] / 100
df[24] = df[20] * df[24] / 100
df[25] = df[20] * df[25] / 100
# names(df)[which(names(df)=="lymphocyte_pct")]  = "lymphocyte_count"
# names(df)[which(names(df)=="monocyte_pct")]    = "monocyte_count"
# names(df)[which(names(df)=="eosinophils_pct")] = "eosinophils_count"
# names(df)[which(names(df)=="basophils_pct")]   = "basophils_count"
# names(df)[which(names(df)=="neutrophils_pct")] = "neutrophils_count"
# categorical variates should be treated as factor.
df[,"edu_cat"]  = factor(df[,"edu_cat"])#,levels=c("< High School","High School","College","College Graduate"))
df[,"race_cat"] = factor(df[,"race_cat"])
df[,"male"]     = factor(df[,"male"])
df[,"smokenow"] = factor(df[,"smokenow"])
```

The recorded explanatory covariates can be divided into 4 groups:

* the first group contains all 18 kinds of pollutants,
* the second group contains all 6 kinds of cells,
* the third group contains all covariates related to "smoking",
* and the last group contains all others remaining.

```{r}
# sub-dataframes.
# sub-dataframes include length and all the pollutants elements
pollutants <- df[,c(1, 2:19)]
# sub-dataframes include length and all the cell elements
cells      <- df[,c(1,20:25)]
# sub-dataframes include length, smoke years, smoke now ,and cotinine
smoke      <- df[,c(1,31:33)]
# ub-dataframes include length, age, gender, education level, race level and BMI
others     <- df[,c(1,26:30)]
if (!(length(pollutants) == 1 + 18)) message("ERROR")
if (!(length(cells)      == 1 + 6))  message("ERROR")
if (!(length(smoke)      == 1 + 3))  message("ERROR")
if (!(length(others)     == 1 + 5))  message("ERROR")
```

### Report Summary Statistics:

```{r Report Summary Statistics}
y.smr = summary(df$length)
```

We found that the mean of the telomere length variable is `r y.smr["Mean"]`
and the range of telomere length variable is `r y.smr["Max."]-y.smr["Min."]`.
The mean and range of the concentrations of the pollutants are summarized below:

```{r}
# get the mean of all the variables in pollutants
p.mean.vec = round(apply(pollutants, 2, FUN = mean)[-1], 4)
# get the range of all the variables in pollutants
p.range.vec = round(apply(pollutants, 2, FUN = function(col) range(col)[2]-range(col)[1])[-1], 4)
p.table = rbind(p.mean.vec, p.range.vec)
rownames(p.table) = c("mean", "range")
p.table
```

The PCB with largest mean is `r colnames(p.table)[c(1:11)][which.max(p.table["mean",c(1:11)])]`.
The PCB with largest range is `r colnames(p.table)[c(1:11)][which.max(p.table["range",c(1:11)])]`.
The dioxin with largest mean is `r colnames(p.table)[c(12:14)][which.max(p.table["mean",c(12:14)])]`.
The dioxin with largest range is `r colnames(p.table)[c(12:14)][which.max(p.table["range",c(12:14)])]`.
The furan with largest mean is `r colnames(p.table)[c(15:18)][which.max(p.table["mean",c(15:18)])]`.
The furan with largest range is `r colnames(p.table)[c(15:18)][which.max(p.table["range",c(15:18)])]`.

The mean of the cells counts are summarized below:

```{r}
# get the mean of all the variables in cells
c.mean.vec = round(apply(cells, 2, FUN = mean)[-1], 4)
# get the range of all the variables in cells
names(c.mean.vec) = sapply(names(df)[20:25], FUN = function(str) substr(str, 1, nchar(str)-nchar("_count")))
c.mean.vec
```

The type of cell with the largest mean is `r names(c.mean.vec)[which.max(c.mean.vec)]`.
The type of cell with the smallest mean is `r names(c.mean.vec)[which.min(c.mean.vec)]`.

```{r}
num_nosmoke <- which(smoke$yrssmoke == min(smoke$yrssmoke))
num_largesmoke <- which(smoke$yrssmoke == max(smoke$yrssmoke))
```
For the smoke subdataframe, the lowest year smoked cigarettes is 0 year and we find that there are `r sum(smoke$yrssmoke==0)` people never smoked.
Dividing by the sample size, we found that there are approximately `r round(sum(smoke$yrssmoke==0)/nrow(smoke)*100)`% people who do not smoke.
The longest years smoked is `r max(smoke$yrssmoke)`.
There was only one person who smoked for this many years.
From the variable "smokenow", we found that
there are `r sum(smoke$smokenow==0)` poeple who do not currently smoke
and there are `r sum(smoke$smokenow==1)` people who currently smoke.
This means that there are currently more than three times as many non-smokers as there are smokers.

```{r}
round(c("BMI mean" = mean(others$BMI), "BMI range" = range(others$BMI)[2] - range(others$BMI)[1],
  "edu_cat 1" = sum(others$edu_cat==1), "edu_cat 2" = sum(others$edu_cat==2),
  "edu_cat 3" = sum(others$edu_cat==3), "edu_cat 4" = sum(others$edu_cat==4),
  "race_cat 1" = sum(others$race_cat==1), "race_cat 2" = sum(others$race_cat==2),
  "race_cat 3" = sum(others$race_cat==3), "race_cat 4" = sum(others$race_cat==4),
  "female" = sum(others$male==0), "male" = sum(others$male==1), "ageyrs mean" = mean(others$ageyrs),
  "ageyrs range" = range(others$ageyrs)[2] - range(others$ageyrs)[1]), 4)
```
```{r fig.width=7,fig.height=3}
par(mfrow = c(1, 2))
plot(others$edu_cat, xlab = "Education", ylab = "Count", main = "Barplot of Education Level")
plot(others$race_cat, xlab = "Race", ylab = "Count", main = "Barplot of Different Race")
```

```{r}
df_cor = df[,-c(27,28,29,32)]
df_cor <- cor(df_cor)
max_cor <- max(df_cor[row(df_cor) != col(df_cor)])
min_cor <- min(df_cor[row(df_cor) != col(df_cor)])
cells_cor <- cor(cells)
cmax_cor <- max(cells_cor[row(cells_cor) != col(cells_cor)])
cmin_cor <- min(cells_cor[row(cells_cor) != col(cells_cor)])
round(c("max cor (df)" = max_cor, "min cor (df)" = min_cor, "max cor (cells)" = cmax_cor,
"min cor (cells)" = cmin_cor), 4)
```
We can see the maximum number of correlation coefficient is `r max(df_cor)` 
which is the relarionship between POP_PCB1 and POP_PCB2.
This means POP_PCB1 and POP_PCB2 have a strong positive relationship.
Also, the minimum number of correlation coefficient is -0.9346368,
which is the strong negative realtionship between eosinophils_pct and lymphocyte_pct.
We also found that length has the smallest correlation coefficient which close to -1 with ageyrs which is -0.4454.
The strongest relationship in the data fram cells is between eosinophils_pct and whitecell_count which is 0.92951238.
This means that eosinophils_pct and whitecell_count has a strong positibe relationship.
The correlation coefficient closest to 1 or -1 between length and other variables in cells dataframe is -0.04949 which is between length and monocyte_pct.
We found that the telomere length does not have strong relationship with any type of cells.

```{r}
pollutants_cor <- cor(pollutants)
pollutants_cor_len <- pollutants_cor[1,]
min_pollutants <- min(pollutants_cor_len[pollutants_cor_len != 1])
```

We can also get that the strongest relationship between length and pollutant is length and POP_PCB2.
However, the correlation coefficient closest to 1 or -1 is -0.2457 which is closer than 0.
This means that the telomere length and POP_PCB2 does not have strong relationship
and the telomere length does not have strong relationship with any pollutant types.



```{r}
##round(c("length and yrssmoke" = cor(smoke[,c(1:2)])[1,2], "length and ln_lbxcot" = cor(smoke[,c(1,4)])[1,2],
##  "yrssmoke and ln_lbxcot" = cor(smoke[,c(2,4)])[1,2]), 4)
```

From the dataframe smoke and others, we can find that all of them do not have a correlation coefficient 
which is close to 1 or -1. This means neither of them have a strong relarionship between each other.
```{r}
##round(c("length and BMI" = cor(others[,c(1:2)])[1,2], "length and ageyrs" = cor(others[,c(1,6)])[1,2],
##  "BMI and ageyrs" = cor(others[,c(2,6)])[1,2]), 4)
```


```{r, fig.width=7}
par(mfrow = c(1, 3))

hist(df$length, breaks = seq(0, 2.5, 0.1),
     xlab = "Telomere Length", 
     main = "Histogram of Telomere Length")
abline(v = mean(df$length), col = "red", lwd = 2)

plot(pollutants$POP_PCB1 ~ pollutants$POP_PCB2, main = "POP_PCB1 VS. POP_PCB2", 
     xlab = "POP_PCB2 (pg/g)", ylab = "POP_PCB1 (pg/g)")
abline(lm(pollutants$POP_PCB1 ~ pollutants$POP_PCB2))

plot(df$length ~ df$ageyrs, main = "Telomere Length VS. Age in Years", 
     xlab = "Age in Years", ylab = "Telomere Length")
abline(lm(df$length ~ df$ageyrs))
```

From the first histogram of the telomere length, we found that it is not really symmetric because it has a longer tail
on the right and there is an outlier which is larger than 2.0.
In the second plot, we can find they have a really strong positive linear relationship.
In the third plot, we can find they have a negative linear relationship between the mean telomere length and age in year.

### Interesting Findings:

* The PCB/dioxin/furan with largest mean and range are the same one.
* We found that the telomere length does not have strong relationship with any of the other variables.
* The relationship between different types of organic pollutants was very strong.
  We found there are strong relationships between different type of organic pollutant PCBs.
  Also, all the correlation coefficient between length and different pollutants are negetive.
* The telomere length and age in year has a negative linear relationship.
  This indicates as people grow up, their telomere length tend to be shorter
  However, they do not have a really strong linear relationship.

### How these inform the rest of our analysis:

We are going to use the variable which has a correlation coefficient that close to 1 or -1 to get a model. 
Also, the variable which has an absolute correlation coefficient close to 1 will have a lower value of VIF
when we do the model selection.
This is because having the absolute correlation coefficient closer to 1 means that the variables have stronger linear relationship.
Thus having lower VIF value also means the variables have stronger relationship.

```{r}
```
